# Importing Data {#importing}

All of the Itrax data is in either text-format or "tagged image format" (`*.tif`). Although this means it is easily read by the various import functions available in `R`, it still needs considerable cleaning and wrangling to get it to a point where it is useable for most analyses. There are three possible approaches to this task:

1. Use existing functions published in the `itraxR` package available from [github.com/tombishop1](https://github.com/tombishop1/itraxR). These are at an early stage and functionality might be broken, but are largely convenience functions for wrangling and analysing Itrax data. 
2. Work in base `R` to wrangle the data. This is perfectly achievable, and much of the current `itraxR` functionality was written this way.

3. Work in the `tidyverse` family of packages and style. For data wrangling tasks, this approach can result in simpler and more resiliant code. 

In this guide examples will be given mostly using the `tidyverse`, and reference will be made to the convenience functions in `itraxR`. 

All of the Itrax data is in either text-format or "tagged image format" (`*.tiff`). Although this means it is easily read by the various import functions available in `R`, it still needs considerable cleaning and wrangling to get it to a point where it is useable. A number of convienience functions are provided in the package `itraxR`.

```{r importing_init, include=FALSE}

```


## Metadata

The scan data file `document.txt` can be parsed using `itraxR::itrax_meta()`. The output is a dataframe from which the individual components can be easily accessed through subsetting functions, for example `as.numeric(itrax_meta()[6:7, 2])` ould return a numeric vector of the start and end position of a scan.

```{r importing_metadata}
library(itraxR)
example(itrax_meta)
```

## XRF Data

### Processed Data

This is the data most commonly used in analysis and it can be quickly imported using `itraxR::itrax_import()`. If you'd like more control over the import process, a typical `Tidyverse` workflow would consist of reading the file (they are tab-delimited), removing variables that are not of interest, and correctly labelling data that should be `NA`. You may also wish to assign a coring depth information at this stage.

```{r importing_processed_flow, echo=FALSE}
library(DiagrammeR)
DiagrammeR(diagram = "graph LR;
    import[read data file] --> select[select variables of interest] 
    select --> na[label bad data as NA];
    na -->|optional| depth[add coring depth/age]
    ",
    height = '100%', width = '100%')
```

### Joining XRF Data

This process can be simplified using `itraxR::itrax_join()`. 

### Raw Data

Sometimes it is useful to work with raw data rather than the calculated intensity data from the `Q-Spec` software. In this case, the raw data can be read directly from the individual files in the relevant directory. For individual measurements this is fairly trivial, although it must be considered that the data output is not calibrated to an energy and the data are in counts, not intensities. If the entire scan is read, some mechanism to iterate through the individual data files, adding them to a structured data object with relevant metadata (positions, for example) is required. 

## Optical Images

In order to make the images useable for plotting beside other data, they need to be cropped to the scanned area and referenced to the `position` or a `depth` variable. Images in `R` are read in as a matrix with 3 dimensions (length, width, and the three colours). 

To begin with the start and end positions of the scan need to be established; this usually involves parsing the metadata file (e.g. `itraxR::itrax_meta()[]`), although in some circumstances they can be obtained from another data source (e.g. `range(myData[,"Position"])`). These then need to be mapped onto individual pixel values using the known size of a single pixel. Once this has been calculated, it is trvial to then subset the part of the image required. 

## Radiographic Images

The workflow for manipulating the processed radiographic images (`*.tif`) is very similar to that for the optical images, the main difference being the matrix only has two dimensions (length and width) as the image is greyscale. However, if there is a desire to manipulate the raw data from the radiographic image, some further work is required because the "pixel" is not square, but rectangular. That is to say the length of the pixel differs from its width. On the core scanner a single pixel has a width across the core of 20 micrometers, but has a variable coverage along the core (usually between 50 and 200 micrometers). The processed image downscales the pixel width to match the pixel length in order to force square pixels, losing some resolution along the way.

## Magnetic Susceptability 