# Importing Data {#importing}

All of the Itrax data is in either text-format or "tagged image format" (`*.tif`). Although this means it is easily read by the various import functions available in `R`, it still needs considerable cleaning and wrangling to get it to a point where it is usable for most analyses. There are three possible approaches to this task:

1. Use existing functions published in the `itraxR` package available from [github.com/tombishop1/itraxR](https://github.com/tombishop1/itraxR). These are at an early stage and functionality might be broken, but are largely convenience functions for wrangling and analysing Itrax data. It is easy to install the package directly from Github using `remotes::install_github("tombishop1/itraxR")`. 
2. Work in base `R` to wrangle the data. This is perfectly achievable, and much of the current `itraxR` functionality was originally written this way.
3. Work in the `tidyverse` family of packages and style. For data wrangling tasks, this approach can result in simpler and more resilient code. 

In this guide examples will be given using the functions provided in `itraxR`, but the processes used in those functions will be explained, and the code used in those functions is fully commented so can be modified to suit particular needs. 

## Metadata

The scan metadata file `document.txt` can be quickly parsed using `itraxR::itrax_meta()`. The output is a dataframe from which the individual components can be easily accessed through subsetting functions; for example `as.numeric(itrax_meta()[6:7, 2])` would return a numeric vector of the start and end position of a scan.  

```{r importing_metadata}
itrax_meta("CD166_19_S1/CD166_19_S1/document.txt")
```

## XRF Data

### Processed Data

This is the data most commonly used in analysis and it can be quickly imported using `itraxR::itrax_import()`. Note that, like for the example data, it is possible to have more than one processed data file. Typically cores have at least two, one created at the time of the scan based on settings for a single point near the top of the sequence, and another from a holistic re-analysis of the sequence. 

```{r proc_data_table}
itrax_import("CD166_19_S1/CD166_19_S1/Results.txt", depth = 0) %>%
  glimpse()
```

### Joining XRF Data

Often a core (sometimes referred to as a drive) is comprised of a sequence of individual sections, which may or may not be overlapping. Often we will want to integrate them into a continuous dataset for analytical purposes. When joining cores that do not overlap, this process is trivial --- the data might simply appended in order of depth, and a new column is added with the identity of the original core section. 

Where overlapping cores are present, there can be multiple measurements at a single depth (on different cores). In these cases not only will the individual measurements need to be re-ordered by depth, but an additional variable should be created that can be used in combination or alone to uniquely identify each measurement. The code below does this by creating an additional variable called `label`, with the name of the original core given in the named list. 

```
mylist <- list(core1 = core1, core2 = core2)
df <- lapply(names(mylist), function(i) within(mylist[[i]], {label <- i})) %>% 
  bind_rows() %>% 
  arrange(depth)
```

This process can be simplified using `itraxR::itrax_join()`, for example:

```{r itrax_join, message=FALSE, warning=FALSE}
# import the core sections
CD166_19_S1 <- itrax_import("CD166_19_S1/CD166_19_S1/Results.txt", depth_top = 0)
CD166_19_S2 <- itrax_import("CD166_19_S2/CD166_19_S2/Results.txt", depth_top = max(CD166_19_S1$depth))
CD166_19_S3 <- itrax_import("CD166_19_S3/CD166_19_S3/Results.txt", depth_top = max(CD166_19_S2$depth))
#join them together
CD166_19 <- itrax_join(list(S1 = CD166_19_S1, S2 = CD166_19_S2, S3 = CD166_19_S3))
rm(CD166_19_S1, CD166_19_S2, CD166_19_S3)
glimpse(CD166_19)
```

### Raw Data

Sometimes it is useful to work with raw data rather than the calculated intensity data from the `Q-Spec` software. In this case, the raw data can be read directly from the individual files in the relevant directory. For individual measurements this is fairly trivial, although it must be considered that the data output is not calibrated to an energy and the data are in counts, not intensities. If the entire scan is read, some mechanism to iterate through the individual data files, adding them to a structured data object with relevant metadata (positions, for example) is required. 

## Optical Images

In order to make the images usable for plotting alongside other data, they need to be cropped to the scanned area and referenced to the `position` or a `depth` variable. The initial crop can be supressed using `trim = FALSE` if the whole image is desired. Images in `R` are read in as a matrix with 3 dimensions (length, width, and the three colours). The data can be imported using `itraxR::itrax_image()`. This function can produce a basic diagram (shown in subsequent section), but here we focus on the structure of the imported object. `$image` is a three dimensional array, the first two dimensions are the length of the core and the width, respectively. Those two dimensions have `"dimnames"` (`rownames()` and `colnames()` respectively). For the `rownames()`, this is an interpolated position, in mm, and for the `colnames()` this is the width in mm, and always begins at 0. The last dimension always has a length of three, and comprises of the red, green and blue values for each of the pixels. `$meta` is a table containing selected data from the scan metadata relevant to the image. 

```{r image_import_0}
itrax_image(file = "CD166_19_S1/CD166_19_S1/optical0.tif",
            meta = "CD166_19_S1/CD166_19_S1/document.txt",
              trim = FALSE) %>% 
  str()
```

## Radiographic Images

The function `itraxR::itrax_radio()` imports the processed radiographic images (`*.tif`) in a very similar way to that for the optical images, the main difference being the matrix only has two dimensions (length and width) as the image is greyscale. 

```{r radio_import_0}
itrax_radiograph(file = "CD166_19_S1/CD166_19_S1_RAD/radiograph0.tif",
                 meta = "CD166_19_S1/CD166_19_S1_RAD/document.txt") %>%
  str()
```

However, if there is a desire to manipulate the raw data from the radiographic image, some further work is required because the "pixel" is not square, but rectangular; that is to say the length of the pixel differs from its width. On the core scanner a single pixel has a width across the core of 20 μm, but has a variable coverage along the core (usually between 50 and 200 μm). The processed image downscales the pixel width to match the pixel length in order to force square pixels, losing some resolution along the way.

In addition, it should be noted that unlike the optical images that always begin from `position == 0`, the radiographic images have defined start and end points just like an XRF scan, the parameters of which can be accessed from the `$meta` object, or using `itraxR::itrax_meta()`.

## Importing Everything

Given the structured nature of individual core section scans, it might be helpful to import everything into a single data object (`list()`). This is particularly helpful when dealing with large numbers of core sections. However, some care should be taken here, as it is easy to end up holding large quantities of data in memory, particularly where images are included. The structure of this can be adapted to your needs, but could look something like the following example. **These are the data that will be used elsewhere in this guide**. 

```{r everything}
CD166_19_S1 <- list(metadata   = itrax_meta("CD166_19_S1/CD166_19_S1/document.txt"),
                    xrf        = itrax_import("CD166_19_S1/CD166_19_S1/Results.txt", 
                                              depth = 0, 
                                              parameters = "all"),
                    image      = itrax_image(file = "CD166_19_S1/CD166_19_S1/optical0.tif",
                                             meta = "CD166_19_S1/CD166_19_S1/document.txt"),
                    radiograph = itrax_radiograph(file = "CD166_19_S1/CD166_19_S1_RAD/radiograph0.tif",
                                                  meta = "CD166_19_S1/CD166_19_S1_RAD/document.txt",
                                                  trim = as.numeric(itrax_meta("CD166_19_S1/CD166_19_S1/document.txt")[6:7,2])))

CD166_19_S2 <- list(metadata   = itrax_meta("CD166_19_S2/CD166_19_S2/document.txt"),
                    xrf        = itrax_import("CD166_19_S2/CD166_19_S2/Results.txt", 
                                              depth = max(CD166_19_S1$xrf$depth), 
                                              parameters = "all"),
                    image      = itrax_image(file = "CD166_19_S2/CD166_19_S2/optical0.tif",
                                             meta = "CD166_19_S2/CD166_19_S2/document.txt"),
                    radiograph = itrax_radiograph(file = "CD166_19_S2/CD166_19_S2_RAD/radiograph0.tif",
                                                  meta = "CD166_19_S2/CD166_19_S2_RAD/document.txt",
                                                  trim = as.numeric(itrax_meta("CD166_19_S2/CD166_19_S2/document.txt")[6:7,2])))

CD166_19_S3 <- list(metadata   = itrax_meta("CD166_19_S3/CD166_19_S3/document.txt"),
                    xrf        = itrax_import("CD166_19_S3/CD166_19_S3/Results.txt", 
                                              depth = max(CD166_19_S2$xrf$depth), 
                                              parameters = "all"),
                    image      = itrax_image(file = "CD166_19_S3/CD166_19_S3/optical0.tif",
                                             meta = "CD166_19_S3/CD166_19_S3/document.txt"),
                    radiograph = itrax_radiograph(file = "CD166_19_S3/CD166_19_S3_RAD/radiograph0.tif",
                                                  meta = "CD166_19_S3/CD166_19_S3_RAD/document.txt",
                                                  trim = as.numeric(itrax_meta("CD166_19_S3/CD166_19_S3/document.txt")[6:7,2])))

CD166_19_xrf <- itrax_join(list(S1 = CD166_19_S1$xrf, S2 = CD166_19_S2$xrf, S3 = CD166_19_S3$xrf))
```
