# Plotting
There are a number of plotting options included in base `R` and some palaeoenvironmental packages like `analouge`, but here we will use `ggplot2` and compatible packages. We will also be using `tidypaleo`, a package made available by [Dewey Dunnington on Github]( https://github.com/paleolimbot/tidypaleo). 

## Plotting XRF Data
For simple biplots, `ggplot2` provides the following solution. This includes different colour coding to differentiate the core sections. 

```{r simpleplot}
ggplot(data = na.omit(CD166_19_xrf), mapping = aes(x = depth, y = `Mo coh`/`Mo inc`)) +
  geom_line(aes(color = label)) + 
  coord_flip() +
  scale_x_reverse() +
  labs(x = "Depth [mm]", color = "Core") +
  theme_classic() +
  theme(legend.position = "none")
```

Where a traditional stratigraphic plot is desired, the `tidypaleo` package provides useful functionality for producing these. However, the data does need to converted to long-form from the existing table. In this way `tidypaleo::facet_geochem_gridh` works in a way similar to `ggplot::facet_wrap` but has been creates a stratigraphic plot. The `elementsList` is used with `select(any_of())` to include only columns that are chemical elements. `Depth` and `label` are also added because they are used in the plot. Although this is useful for producing large-format summary diagrams of the data, these do not render well in smaller plot areas, so a small subset of variables has been selected manually; see the code comments below. Note that in the line `mutate()`, the vector passed to `levels = ` controls the order in which the plots appear. In this example, they will appear in order of atomic weight, and the `coh/inc` ratio will always be last. 

```{r palaeoplot}
xrfStrat <- CD166_19_xrf %>% 
  mutate(`coh/inc` = `Mo coh`/`Mo inc`) %>%
  select(any_of(elementsList), `coh/inc`, depth, label) %>% # this is useful but produces a very large plot that doesn't render well
  select(Fe, Ti, `coh/inc`, Mn, depth, label) %>% # a smaller set of elements is defined here manually.
  tidyr::pivot_longer(!c("depth", "label"), names_to = "elements", values_to = "peakarea") %>% 
  tidyr::drop_na() %>%
  mutate(elements = factor(elements, levels = c(elementsList, "coh/inc"))) %>% # note that the levels controls the order

  ggplot(aes(x = peakarea, y = depth)) +
    tidypaleo::geom_lineh(aes(color = label)) +
    scale_y_reverse() +
    scale_x_continuous(n.breaks = 2) +
    facet_geochem_gridh(vars(elements)) +
    labs(x = "peak area", y = "Depth [mm]") +
    tidypaleo::theme_paleo() +
    theme(legend.position = "none")

print(xrfStrat)
```

Notice that in the previous example the data must be converted to "long-form" from "short-form" for `tidypaleo`, using `tidyr::pivot_longer`. "Short-form" data has observations as rows, and variables as columns, whereas "long-form" data has a column for `names`, which in this case are the elements, and a column for `values`, in this case the peak intensities. This results in many rows, but few columns. 

```{r longform}
CD166_19_xrf %>% 
  select(any_of(elementsList), depth, label) %>%
  tidyr::pivot_longer(!c("depth", "label"), names_to = "elements", values_to = "peakarea") %>%
  glimpse()
```

## Plotting Images
The package `itraxR` provides functions for importing and creating basic plots of both the photographic and radiographic images. For the optical images, it is almost always necessary to trim the image to the extent of the actual scan, or extraneous image is included (e.g. the calibration card in the example shown below). The function `itraxR::itrax_image()` allows the extent of the scan to be defined, or extracted from the metadata. In the example below, `trim = FALSE` forces the whole image to be shown. When dealing with high aspect ratio images like these, it is often useful to use `svgPanZoom::svgPanZoom()` instead of `print()` to view the output.

```{r plot_image}
myImage <- itrax_image(file = "CD166_19_S1/CD166_19_S1/optical0.tif",
                       meta = "CD166_19_S1/CD166_19_S1/document.txt",
                       trim = FALSE,
                       plot = TRUE
                       )
```

For radiographic images the function `itraxR::itrax_radiograph()` can be used using an identical syntax as `itraxR::itrax_image()`. Often the contrast needs to be adjusted, so a contrast-adjusted radiograph is supplied in the example data (`radiograph0.tif`). 

```{r itrax_image}
myRadio <- itrax_radiograph(file = "CD166_19_S1/CD166_19_S1_RAD/radiograph0.tif",
                            meta = "CD166_19_S1/CD166_19_S1_RAD/document.txt",
                            plot = TRUE
                            )
```

The simple plots generated as a side effect of these functions are sometimes sufficient for simple plots, but for more complex diagrams you will need to generate them from scratch. For example, it is often desirable to overplot the radiograph on the optical image. The code below does this. Note the calculations made for the `xmin = ` and `ymin = ` parameters for `annotation_custom()` when adding the radiograph `rasterGrob()`; these correctly position the radiograph along the centre of the image.

```{r overplot}
ggplot() +
  scale_y_continuous(limits = rev(range(as.numeric(row.names(myImage$image))))) +
  scale_x_continuous(breaks = round(range(as.numeric(colnames(myImage$image))), 0),
                     limits = range(as.numeric(colnames(myImage$image)))) +
  labs(y = "Position [mm]", x = "[mm]") +
  annotation_custom(rasterGrob(myImage$image,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    ymax = max(as.numeric(row.names(myImage$image))),
                    ymin = min(as.numeric(row.names(myImage$image))),
                    xmin = min(as.numeric(colnames(myImage$image))),
                    xmax = max(as.numeric(colnames(myImage$image)))
  ) +
  annotation_custom(rasterGrob(myRadio$image,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    ymax = max(as.numeric(row.names(myRadio$image))),
                    ymin = min(as.numeric(row.names(myRadio$image))),
                    xmin = mean(as.numeric(colnames(myImage$image))) - mean(as.numeric(colnames(myRadio$image))),
                    xmax = mean(as.numeric(colnames(myImage$image))) + mean(as.numeric(colnames(myRadio$image)))
  ) +
  coord_fixed(ratio = 1)
```

```{r tidyup2, include=FALSE}
rm(myImage, myRadio)
```

It is often desirable to combine a sequence of cores. This is trivial for cores like `CD166_19` where there are no overlapping sections. However, note that the coring depth information is different from the `position` that these images are referenced to. The coring depth is not contained in any of the Itrax metadata, and should be recorded elsewhere, whereas `position` refers to the coordinates on the Itrax core holder. In the code below the depth information is extracted from the combined XRF data created in a previous chapter (for example using `max(CD166_19[which(CD166_19$label == "S2"),]$depth)` to get the maximum depth of the middle section. The radiographs can be added in the same fashion as the previous example.

```{r overplot_sequence}
ggplot() +
  scale_y_continuous(limits = rev(range(CD166_19_xrf$depth))) +
  scale_x_continuous(breaks = round(range(as.numeric(colnames(CD166_19_S1$image$image))), 0),
                     limits = range(as.numeric(colnames(CD166_19_S1$image$image)))) +
  coord_fixed(ratio = 1) +
  labs(y = "Depth [mm]", x = "[mm]") +
  annotation_custom(rasterGrob(CD166_19_S1$image$image,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    ymax = max(CD166_19_S1$xrf$depth),
                    ymin = min(CD166_19_S1$xrf$depth),
                    xmin = min(as.numeric(colnames(CD166_19_S1$image$image))),
                    xmax = max(as.numeric(colnames(CD166_19_S1$image$image)))
  ) +
  annotation_custom(rasterGrob(CD166_19_S2$image$image,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    ymax = max(CD166_19_S2$xrf$depth),
                    ymin = min(CD166_19_S2$xrf$depth),
                    xmin = min(as.numeric(colnames(CD166_19_S2$image$image))),
                    xmax = max(as.numeric(colnames(CD166_19_S2$image$image)))
  ) +
  annotation_custom(rasterGrob(CD166_19_S3$image$image,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    ymax = max(CD166_19_S3$xrf$depth),
                    ymin = min(CD166_19_S3$xrf$depth),
                    xmin = min(as.numeric(colnames(CD166_19_S3$image$image))),
                    xmax = max(as.numeric(colnames(CD166_19_S3$image$image)))
  )
```

Where there are overlaps it is often desirable to plot the cores adjacent to one another. It is simply a case of adjusting the `xmin = ` and `xmax = ` parameters in the `annotation_custom()` function for every other section. In this context the x-axis labels become largely meaningless and can be omitted. 

```{r sideplot_sequence}
ggplot() +
  scale_y_continuous(limits = rev(range(CD166_19_xrf$depth))) +
  scale_x_continuous(breaks = round(c(0, max(as.numeric(colnames(CD166_19_S2$image$image)))*2)),
                     limits = c(0, max(as.numeric(colnames(CD166_19_S2$image$image)))*2)) +
  coord_fixed(ratio = 1) +
  labs(y = "Depth [mm]", x = "[mm]") +
  annotation_custom(rasterGrob(CD166_19_S1$image$image,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    ymax = max(CD166_19_S1$xrf$depth),
                    ymin = min(CD166_19_S1$xrf$depth),
                    xmin = min(as.numeric(colnames(CD166_19_S1$image$image))),
                    xmax = max(as.numeric(colnames(CD166_19_S1$image$image)))
  ) +
  annotation_custom(rasterGrob(CD166_19_S2$image$image,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    ymax = max(CD166_19_S2$xrf$depth),
                    ymin = min(CD166_19_S2$xrf$depth),
                    xmin = max(as.numeric(colnames(CD166_19_S2$image$image))),
                    xmax = max(as.numeric(colnames(CD166_19_S2$image$image)))*2
  ) +
  annotation_custom(rasterGrob(CD166_19_S3$image$image,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    ymax = max(CD166_19_S3$xrf$depth),
                    ymin = min(CD166_19_S3$xrf$depth),
                    xmin = min(as.numeric(colnames(CD166_19_S3$image$image))),
                    xmax = max(as.numeric(colnames(CD166_19_S3$image$image)))
  )
```

Note that omitting the line `coord_fixed(ratio = 1)` allows the aspect ratio of the plot to be reshaped. Although usually it is desirable to keep the plot in the correct shape, sometimes for very long sequences the images end up so narrow as to be quite useless, so the x-axis can be stretched to accentuate the features in it. The example below simply omits the `coord_fixed()` line from the previous example.

```{r sideplot_sequence_aspectratio, echo=FALSE}
ggplot() +
  scale_y_continuous(limits = rev(range(CD166_19_xrf$depth))) +
  scale_x_continuous(breaks = round(c(0, max(as.numeric(colnames(CD166_19_S2$image$image)))*2)),
                     limits = c(0, max(as.numeric(colnames(CD166_19_S2$image$image)))*2)) +
  #coord_fixed(ratio = 1) +
  labs(y = "Depth [mm]", x = "[mm]") +
  annotation_custom(rasterGrob(CD166_19_S1$image$image,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    ymax = max(CD166_19_S1$xrf$depth),
                    ymin = min(CD166_19_S1$xrf$depth),
                    xmin = min(as.numeric(colnames(CD166_19_S1$image$image))),
                    xmax = max(as.numeric(colnames(CD166_19_S1$image$image)))
  ) +
  annotation_custom(rasterGrob(CD166_19_S2$image$image,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    ymax = max(CD166_19_S2$xrf$depth),
                    ymin = min(CD166_19_S2$xrf$depth),
                    xmin = max(as.numeric(colnames(CD166_19_S2$image$image))),
                    xmax = max(as.numeric(colnames(CD166_19_S2$image$image)))*2
  ) +
  annotation_custom(rasterGrob(CD166_19_S3$image$image,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    ymax = max(CD166_19_S3$xrf$depth),
                    ymin = min(CD166_19_S3$xrf$depth),
                    xmin = min(as.numeric(colnames(CD166_19_S3$image$image))),
                    xmax = max(as.numeric(colnames(CD166_19_S3$image$image)))
  )
```

For lengthy sequences, the repetition of the `annotation_custom()` block of code is inefficient, but it can be easily placed inside a function. The functions `ggItraxImage()` and `ggItraxRadio()` shown below allow this, but they might need small modifications depending on your data structure and the desired output. 

```{r plotting_functions}
ggItraxImage <- function(section, xposition){
  
  if(xposition == "l"){
  annotation_custom(rasterGrob(section$image$image,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    ymax = max(section$xrf$depth),
                    ymin = min(section$xrf$depth),
                    xmin = min(as.numeric(colnames(section$image$image))),
                    xmax = max(as.numeric(colnames(section$image$image)))
  )} else if(xposition == "r"){
    annotation_custom(rasterGrob(section$image$image,
                                 width = unit(1, "npc"),
                                 height = unit(1, "npc")),
                      ymax = max(section$xrf$depth),
                      ymin = min(section$xrf$depth),
                      xmin = max(as.numeric(colnames(section$image$image))),
                      xmax = max(as.numeric(colnames(section$image$image)))*2)
  } else{stop("Specify `l`eft or `r`ight for the image.")}
  
}

ggItraxRadio <- function(section, xposition){
  if(xposition == "l"){
    annotation_custom(rasterGrob(section$radiograph$image,
                                 width = unit(1, "npc"),
                                 height = unit(1, "npc")),
                      ymin = min(section$xrf$depth),
                      ymax = max(section$xrf$depth),
                      xmin = mean(as.numeric(colnames(section$image$image))) -
                        mean(as.numeric(colnames(section$radiograph$image))),
                      xmax = mean(as.numeric(colnames(section$image$image))) +
                        mean(as.numeric(colnames(section$radiograph$image)))
    )
  } else if(xposition == "r"){
    annotation_custom(rasterGrob(section$radiograph$image,
                                 width = unit(1, "npc"),
                                 height = unit(1, "npc")),
                      ymin = min(section$xrf$depth),
                      ymax = max(section$xrf$depth),
                      xmin = max(as.numeric(colnames(section$image$image))) +
                        mean(as.numeric(colnames(section$image$image))) -
                        mean(as.numeric(colnames(section$radiograph$image))),
                      xmax = max(as.numeric(colnames(section$image$image))) +
                        mean(as.numeric(colnames(section$image$image))) +
                        mean(as.numeric(colnames(section$radiograph$image)))
    )
  } else(stop("Specify `l`eft or `r`ight for the radiograph."))
}
```

Recall that `ggplot` objects can be saved and recalled rather than immediately printed. Let us take the plot generated in a previous example, but assign it using `imagePlot <- ggplot() + ...`, whilst also using our newly created functions to shorten the code.

```{r imageplot}
imagePlot <- ggplot() +
  scale_y_continuous(limits = rev(range(CD166_19_xrf$depth))) +
  scale_x_continuous(breaks = round(c(0, max(as.numeric(colnames(CD166_19_S1$image$image)))*2)),
                     limits = c(0, max(as.numeric(colnames(CD166_19_S1$image$image)))*2)) +
  labs(y = "Depth [mm]", x = "[mm]") +
  ggItraxImage(CD166_19_S1, "l") +
  ggItraxImage(CD166_19_S2, "r") +
  ggItraxImage(CD166_19_S3, "l")
```

The radiographs can be added afterwards without starting from scratch. We can also modify plot parameters, for example, the x-axis labels can be removed. 

```{r add_radiographs}
imagePlot <- imagePlot +
  ggItraxRadio(CD166_19_S1, "l") +
  ggItraxRadio(CD166_19_S2, "r") +
  ggItraxRadio(CD166_19_S3, "l") +
    theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

print(imagePlot + coord_fixed(ratio = 1))
```

Finally, it may be necessary to plot the images horizontally rather than the more traditional vertical plots often used in down-core data presentation. `aperm()` and `t()` may be used for the optical and radiographic images respectively to transpose the x- and y-coordinates, but the images will need to be mirrored (using `[]`) so that the tops of the sections are on the right-hand side and the graphs run from oldest to youngest when read left-to-right, in the generally accepted order of these types of graph. For the example below, this is achieved in the first and second lines for the optical and radiographic images respectively. 

```{r sideways_plot}
mySidewaysImage <- aperm(CD166_19_S1$image$image, c(2,1,3))[, dim(aperm(CD166_19_S1$image$image, c(2,1,3)))[2]:1, ]
mySidewaysRadio <- t(CD166_19_S1$radiograph$image)[, dim(t(CD166_19_S1$radiograph$image))[2]:1]

ggplot() +
  scale_x_continuous(limits = rev(range(as.numeric(colnames(mySidewaysImage))))) +
  scale_y_continuous(breaks = round(range(as.numeric(row.names(mySidewaysImage))), 0),
                     limits = range(as.numeric(row.names(mySidewaysImage)))) +
  labs(y = "[mm]", x = "Position [mm]") +
  annotation_custom(rasterGrob(mySidewaysImage,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    xmax = max(as.numeric(colnames(mySidewaysImage))),
                    xmin = min(as.numeric(colnames(mySidewaysImage))),
                    ymin = min(as.numeric(row.names(mySidewaysImage))),
                    ymax = max(as.numeric(row.names(mySidewaysImage)))
  ) +
  annotation_custom(rasterGrob(mySidewaysRadio,
                               width = unit(1, "npc"),
                               height = unit(1, "npc")),
                    xmax = max(as.numeric(colnames(mySidewaysRadio))),
                    xmin = min(as.numeric(colnames(mySidewaysRadio))),
                    ymin = mean(as.numeric(row.names(mySidewaysImage))) - mean(as.numeric(row.names(mySidewaysRadio))),
                    ymax = mean(as.numeric(row.names(mySidewaysImage))) + mean(as.numeric(row.names(mySidewaysRadio)))
  ) +
  coord_fixed(ratio = 1)

```

```{r cleanup01, include=FALSE}
rm(mySidewaysImage, mySidewaysRadio)
```

## Combining Images with XRF Data
It is possible to combine different plots. For example, it is often desirable to plot the XRF data alongside the core imagery. By not passing `coord_fixed(ratio = 1)` to the image plot, its relative width can be controlled via `widths = `, which makes for a visually pleasing plot. Note also that the y-axis labels have been removed from all but one of the plots to avoid duplication. 

```{r combining_images}
egg::ggarrange(imagePlot + theme_paleo(), 
               xrfStrat + theme(axis.title.y = element_blank(),
                                axis.text.y  = element_blank(),
                                axis.ticks.y = element_blank()), 
               ncol = 2, 
               widths = c(1, 4) # these are relative. For c(1, 5), the first plot will be 1/5th the width of the second.
               )
```
