# Transforming Data
The XRF data typically reported from the Itrax core scanner come from the spectral processing software Q-Spec. The output is usually in the form of an intensity, which is a dimensionless metric derived from the size of the spectral peak for a particular element, above the  background Bremsstrahlung radiation, sometimes normalised for the tube current and/or counting time. 

## Ratios and Normalisation
These data are compositional, and represent the changes in the relative proportions of all components of the matrix, measured and un-measured. As such it is likely the data will need transforming for certain types of multivariate analysis. As previously mentioned, these data are dimensionless, and as such do not represent a quantity, but are directly related to the absolute amount of a particular element in the matrix. It is often the case that ratios of elements are used to represent changes in composition --- this is sometimes referred to as normalisation, or normalising one element against another. 

For example, particular element ratios can be used to make some environmental inference. Log-ratios are usually preferable here as they are resilient to matrix effects. In this context, the Ca/Ti log-ratio is used to infer a relative measure of productivity (that it, it can identify the hemipelagic sediments in the sequence). Note the use of `abs()` has the effect of ensuring that switching the denominator and numerator has no effect.

```{r}
CD166_19_xrf %>%
  mutate(`log(Fe/Ca)` = abs(log(Fe/Ca))) %>%
  filter(qc == TRUE) %>%
  
  ggplot(aes(x = depth, y = `log(Fe/Ca)`)) +
  geom_line() + 
  scale_x_reverse()
```
Sometimes z-scores are useful to use when units don't have meaningful units. They scale the data in standard deviations from their mean, so give a useful quantity for the magnitude of change. For example:

```{r}
CD166_19_xrf %>%
  mutate(`log(Fe/Ca)` = abs(log(Fe/Ca))) %>%
  mutate(zscore = (`log(Fe/Ca)` - mean(`log(Fe/Ca)`, na.rm = TRUE))/sd(`log(Fe/Ca)`, na.rm = TRUE)) %>%
  filter(qc == TRUE) %>%
  
  ggplot(aes(x = depth, y = zscore)) +
  geom_line() + 
  scale_x_reverse() +
  ylab("log(Fe/Ca) [z-score]") 
```

It is trivial to calculate element ratios, to the extent that these can often simply be calculated where they are required rather than saving them to memory. For example, if a plot of the Compton divided by the Rayleigh scatter was desired, there is no need to save the computed value to a new variable (e.g. `coh_inc <- df$Mo.coh/df$Mo.inc`) --- simply define the calculation during plotting. To calculate ratios for all elements at once, use `mutate(across(any_ofelementsList))`, where `elementsList` is a list of chemical elements extracted from `data(PeriodicTable)`.

```
CD166_19_xrf %>% 
  mutate(across(any_of(elementsList)) /`Mo inc`)
```

This can be integrated into the work flow from the data tidying chapter, for example:

```{r}
CD166_19_xrf %>% 
  # transform
  mutate(across(any_of(elementsList)) /`Mo inc`) %>%
  
  # identify acceptable observations
  filter(validity == TRUE) %>%
  
  # identify acceptable variables
  select(any_of(myElements), depth, label) %>% 
  
  # pivot
  tidyr::pivot_longer(!c("depth", "label"), names_to = "elements", values_to = "peakarea") %>% 
  mutate(elements = factor(elements, levels = c(elementsList, "coh/inc"))) %>%
  
  # plot
  ggplot(aes(x = peakarea, y = depth)) +
    tidypaleo::geom_lineh(aes(color = label)) +
    scale_y_reverse() +
    scale_x_continuous(n.breaks = 2) +
    facet_geochem_gridh(vars(elements)) +
    labs(x = "peak area / Mo. inc.", y = "Depth [mm]") +
    tidypaleo::theme_paleo() +
    theme(legend.position = "none")
```

Note that where zero values are encountered in divisions, dividing by zero will lead to `Inf` values. This can cause issues with plotting data, and these values should be cleaned into `NA` before plotting. For example:

```{r message=FALSE, warning=FALSE}
ggarrange(
  
  CD166_19_xrf %>%
  mutate(`Fe/Sc` = Fe/Sc) %>%
  ggplot(aes(x = depth, y = `Fe/Sc`)) + 
  geom_line() + 
    scale_x_reverse() + 
    ggtitle("w/ Inf."),
  
  CD166_19_xrf %>%
  mutate(`Fe/Sc` = Fe/Sc) %>%
  mutate_if(is.numeric, list(~na_if(., Inf))) %>% # convert all Inf to NA
  ggplot(aes(x = depth, y = `Fe/Sc`)) + 
  geom_line() + 
    scale_x_reverse() + 
    ggtitle("na_if(., Inf)")
  
)
```

## Running Mean and Other Window Functions

Where a signal is noisy but appears to exhibit some signal it may be appropriate to use a running mean to "smooth" the signal. However, considerable caution should be exercised in the use of this tool. It is rare for an analysis to be genuinely improved by the use of running means, although it can artificially improve statistics for some tests. When visualising data using a running mean the original, unmodified data should always be shown alongside to avoid any misunderstanding. This method can be used for any suitable window function (e.g. `min()`, `max()`, `range()` and `sd`.)

```{r warning=FALSE}
CD166_19_xrf %>%
  # uses a 50 point running mean (50 mm for this data); 25 before, 25 after
  mutate(across(any_of(elementsList), 
                function(x){unlist(slider::slide(x, mean, .before = 25, .after = 25))}
                )
         ) %>%
  ggplot(mapping = aes(x = depth, y = Ca)) + 
  geom_line(data = CD166_19_xrf, col = "grey80") + 
  geom_line() + 
  scale_x_reverse() +
  theme_paleo()
```

To plot the running means in a stratigraphic diagram, the smoothed data has to be labelled and combined with the original data so it can be faceted.

```{r message=FALSE}
# make the xrf plot with running means
full_join(y = CD166_19_xrf %>%
                        as_tibble() %>%
                        # uses a 50 point running mean (50 mm for this data); 25 before, 25 after
                        mutate(across(any_of(c(elementsList)), 
                                      function(x){unlist(slider::slide(x, mean, .before = 25, .after = 25))}
                                      )
                               ) %>%
                        mutate(type = "mean"), 
                      x = CD166_19_xrf %>% 
                        as_tibble() %>% 
                        mutate(type = "raw")
                      ) %>% 
  filter(validity == TRUE) %>%
  select(Fe, Ti, Cu, Pb, Si, MSE, Mn, depth, label, type) %>%
  tidyr::pivot_longer(!c("depth", "label", "type"), names_to = "elements", values_to = "peakarea") %>% 
  tidyr::drop_na() %>%
  mutate(elements = factor(elements, levels = c("MSE", elementsList))) %>%
  mutate(label = as_factor(label),
         type = as_factor(type)
         ) %>%
  
  glimpse() %>%
  
  ggplot(aes(x = peakarea, y = depth)) +
  tidypaleo::geom_lineh(aes(group = type, colour = label, alpha = type)) +
  scale_alpha_manual(values = c(0.1, 1)) +
  scale_y_reverse() +
  scale_x_continuous(n.breaks = 2) +
  facet_geochem_gridh(vars(elements)) +
  labs(x = "peak area", y = "Depth [mm]") +
  tidypaleo::theme_paleo() +
  theme(legend.position = "none",
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```

