--- 
title: "Using Itrax Data in R"
author: "Thomas Bishop"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: biblio.bib
biblio-style: apalike
link-citations: yes
description: This book is a guide to importing, tidying, manipulating and analysing
  Itrax core scanner data in R.
---

```{r setup, include=FALSE, echo=FALSE}
pacman::p_load("tidyverse" , "PeriodicTable", 
               "egg", "errors", 
               "forecast"  , "slider",
               "compositions", "ggpmisc",
               "itraxR", "tidypaleo",
               "ordr", "knitr",
               install = FALSE)
               
opts_knit$set(root.dir = "CD166_19_(2020)/")
```

# Preface {-}

Itrax core scanners (manufactured by Cox Analytical Systems, Sweden) are used in palaeoceanography, palaeolimnological, geological and other "down-core" studies of sedimentary core material. The multi-sensory data can include radiography, optical images, magnetic susceptibility, but most importantly energy-dispersive x-ray florescence (ED-XRF) measurements of elemental abundance. The data can be harder to work with compared to some other palaeoenvironmental techniques because:

1. Very large quantities of data are produced. ED-XRF measurements can be made every 100-200 μm, so for long core sections, these datasets are large. Simple line-graphs can become problematic, and multivariate analysis can become unworkable on some software.
2. Images can be very large and need correct alignment. Combining them with line-graphs or other images can be troublesome. 
3. ED-XRF elemental data is compositional, but dimensionless (they do not have units attached e.g. [ppm]). This can make the use of traditional statistical tools and tests problematic. 

This guide comes from a series of seminars offered to users of the Itrax core scanner at The University of Manchester Geography Laboratories in 2020.

## The itraxR Package
The `itraxR` package offers a range of convenience functions for working with Itrax data. The book mostly uses the functions in this package, but the source code is available and fully commented, so is easy to as a basis for other work. The package and code are available from [github.com/tombishop1/itraxR](https://github.com/tombishop1/itraxR). The package is available via CRAN, so can be installed using `install.packages("itraxR")`, but the CRAN repository is updated less frequently. *This guide always uses the latest version available on the Github repo.*

```{r}
packageVersion("itraxR")
```

## Prerequisites
This guide assumes a basic knowledge of `R` and the `tidyverse`, including data types, assignments, and pipes. It also assumes a background knowledge of the core scanner and the nature of the data it produces; see @Croudace2019 and references therein. Some of the sections on data analysis assume some knowledge of compositional data analysis. 

## Example Data
All of these examples given in this book are from Piston core CD166/19, which was recovered from the Agadir Basin during the RRS Charles Darwin expedition CD166 in 2004 (@Wynn2005). The core site is located at 31°31’ N, 17°11.77’ W at a water depth of 4502 m. The coring operation recovered 4.3 m of sediment composed of hemipelagite and turbidites. The hemipelagic sediments range from cream-brown carbonate-rich marl/ooze to red-brown clays. The turbidites include siliciclastic, volcaniclastic and calcareous sediments.

The split core sections were analysed using the Cox Analytical Systems Itrax core scanner at the British Ocean Sediment Core Research Facility, National Oceanography Centre (Southampton, UK). The split core surfaces were cleaned and loaded to the core scanner along with a radiographic reference sample (@Francus2015) and a colour card. The cores were first scanned to measure the sample surface height and to obtain optical images. The core sections were then covered with Mylar film to reduce sample desiccation during XRF scanning and x-radiography.

XRF data were acquired using a Molybdenum X-ray tube set at 30 kV and 30 mA with a dwell time of 30 seconds and a step size of 1 mm. X-radiographs were acquired with voltage and current set to 55 kV and 50 mA, respectively, and dwell time set to 500 ms. Step size for the x-radiography was set to 200 µm. Two replicate XRF scans were also performed for each of the three core sections that make up this core. The XRF settings for the replicates were kept consistent with those of the main XRF scan. The interval selected for the replicate scans ranged from 10 cm to 15 cm of the the total sample length. All raw XRF raw data were reprocessed using the QSpec spectral analysis software to optimise peak fitting. 

The data can be downloaded from the Github pages that form the source for this document, available at [github.com/tombishop1/itraxBook](https://github.com/tombishop1/itraxBook). They are also available in the same form from the [PANGAEA database](https://doi.pangaea.de/10.1594/PANGAEA.955347) (@Bishop2023).

## Utility Data

It is often useful to have a list of symbols used to represent elements for various subsetting functions. This can be extracted from the example data included with the package `periodicTable` as follows. The `vector` called `elementslist` will be referred to elsewhere in this book. 

```{r tableofelements}
data(periodicTable)
elementsList <- periodicTable$symb
rm(periodicTable)
```

## Citing This Work 

`R` and `itraxR` can be cited as follows:

```{r}
citation()
citation("itraxR")
```

You should also cite other packages you use, for example `dplyr` and `tidypaleo`. 

This book can be cited as per the citation information in the Github repository in which it is located: [github.com/tombishop1/itraxBook](https://github.com/tombishop1/itraxBook). 
